---
title: "Exploration of UN Data Using tidyverse"
output: html_notebook
---

```{r}
library(tidyverse)
```


1.	Using the `read_csv()` function, read the GDP dataset into your notebook as a tibble called `gdp_df`. After reading it in, inspect the first 10 rows and then inspect the last 10 rows. 

```{r}
gdp_df <- read.csv("../data/gdp_per_capita.csv")
```

```{r}
gdp_df |>
  head(10)
```

```{r}
gdp_df |> 
  tail(10)
```

2. Drop the 'Value Footnotes' column, and rename the remaining columns to 'Country', 'Year', and 'GDP_Per_Capita'.

```{r}
gdp_df <- gdp_df |> 
  select(!Value.Footnotes) |> 
  rename(Country = Country.or.Area, Gdp_Per_Capita = Value)
```

3. Which years are represented in this dataset? How many observations are there per year? Make a plot to view the number of observations per year.

```{r}
gdp_df |> 
  group_by(Year) |> 
  count()
```

```{r}
gdp_df |> 
  ggplot(aes(x = Year)) +
  geom_bar() +
  labs(title = "Number of Observations by Year")
```

4. How many countries are represented in this dataset? Which countries have the fewest observations?

```{r}
gdp_df |> 
  summarize(num_countries = n_distinct(Country))
```

```{r}
gdp_df |> 
  group_by(Country) |> 
  count()
```

5. Create a new dataframe by subsetting `gdp_df` to just the year 2014. Call this new dataframe `gdp_2014`.

```{r}
gdp_2014 <- gdp_df |> 
  filter(Year == 2014)
```



